### 2018-05-18 2_wmm.tex
Consider minor changes.
Consider Professor's comments on abstract.

    \item Events from the same process are related by the \textit{scope relation} \sr{}~$\subseteq~\mathbb{E}~\times~\mathbb{E}$.
    In contrast to the \tool{herd} tool, the \porthos[2] does not use hierarchy of scopes (depicted as the scope tree); instead, it uses simple labels that indicate which process has produced certain event.
\end{itemize}

@@

For instance, the \fr-relation is defined as a sequence of inverted \rf-relation and \co-relation: $\texttt{fr} = (\invrel{rf};\texttt{co})$. As an example memory model definition in \cat{} language, the the x86-TSO can be found in Appendix~\ref{apx:x86cat}. This memory model asserts acyclicity of communication relation, \texttt{po-loc} -relation, \texttt{mfence}-relation and some other derived relations~\cite{owens2009better}.

>>>>

    \item Events from the same process are related by the \textit{scope relation} \sr{}~$\subseteq~\mathbb{E}~\times~\mathbb{E}$.
    In contrast to the \tool{herd} tool, \porthos[2] does not use hierarchy of scopes (depicted as the scope tree); instead, it uses simple labels that indicate which process has produced certain event.
\end{itemize}

@@

For instance, the \fr-relation is defined as a sequence of inverted \rf-relation and \co-relation: $\texttt{fr} = (\invrel{rf};\texttt{co})$.
As an example of memory model definition in \cat{} language, the x86-TSO model~\cite{herd10tutorial} can be found in Appendix~\ref{apx:x86cat}.
This memory model asserts acyclicity of communication relation, \texttt{po-loc} -relation, \texttt{mfence}-relation and some other derived relations~\cite{owens2009better}.

### 2018-05-18 3_encod.tex
Consider minor changes.
Consider Professor's comments on abstract.

The other idea is to use satisfiability solvers for symbolic exploration of state space~\cite{clarke2001bounded}.
In this approach, the state space exploration consists of the sequence of queries to the SAT-solver, represented as boolean formulas that encode the constraints of the model and the finite path to a state in the corresponding transition system.  
%This approach uses an iterative process of constructing queries to the SAT-solver as a boolean formula which encodes the constraints of the model and the finite path to a state in the corresponding transition system. 
%Due to the SAT-solver.
This technique is called \textit{bounded model checking (BMC)} as the search process is being repeated up to the user-defined bound $k$, which may result to incomplete analysis in general case.
However, there exist numerous techniques for making BMC complete for finite-state systems~(e.g.,~\cite{shtrichman2000tuning}).
%As well as the BMC problem, the approach used by the \porthos{}

@@

The control-flow of a program is represented by the \textit{control-flow graph}, a directed acyclic connected graph with single source and multiple sink nodes. %TODO: footnote why multiple sinks
In a control-flow graph, there are two types of edges (transitions): \textit{primary transitions} that denote unconditional jumps or if-true-transitions (pictured with solid lines), and \textit{alternative transitions} that denote if-false-transitions (pictured with dotted lines).
Each node on graph can have either one successor (primary) or two successors (both primary and alternative); only computation events can serve as a branching point).
However, each merge node can have any positive number of predecessors, where each edge may be either primary or alternative.

While working on the \porthos[2], we have applied some modifications to the encoding scheme for the control-flow.
The changes are conditioned by the need to be able to process an arbitrary control-flow produced by conditional and unconditional jumps of the C language.
For that, we compile the \textit{Abstract Syntax Tree~(AST)} of the parsed C-code to the plain event-flow graph.
We show %TODO! to show, actually!!!
that the new encoding is smaller than the old one used in \porthos{} since it does not produces new variables for each high-level statement of the input language.
%The \tool{PORTHOS} tool encodes the \textit{instructions}, which have recursive nature, in a recursive manner. This means, it adds synthetic composite instructions for both linear (sequential) and non-linear (branching) instructions. 
For instance, \porthos{} used the encoding scheme where the control-flow of the sequential instruction $i_1 = i_2; i_3$ was encoded as
$\phi_{CF}(i_2;i_3) \defeq (cf_{i_1} \Leftrightarrow (cf_{i_2} \land cf_{i_3})) \land \phi_{CF}(i_2) \land \phi_{CF}(i_3)$,
and control-flow of the branching instruction $i_1 = (c\,?\,i_2\!:\!i_3)$ was encoded as
$\phi_{CF}(c\,?\,i_2\!:\!i_3) \defeq (\textit{cf}_{i_1} \Leftrightarrow (\textit{cf}_{i_2} \lor \textit{cf}_{i_3})) \land \phi_{CF}(i_2) \land \phi_{CF}(i_3)$
(here we used the notation of C-like ternary operator `\texttt{x?y:z}' for defining the conditional expression `$\texttt{if}\;x\;\texttt{then}\;y\;\texttt{else}\;z$').
In contrast, the new scheme implemented in \porthos[2] firstly compiles the recursive high-level code into the linear low-level event-based representation, that is then encoded into an SMT-formula. The encoding of branching nodes depends on the \textit{guards}, the value of conditional variable on the branching state, which in turn is encoded as data-flow constraint (see Section~\ref{ch:enc:enc:df}).

@@

\begin{align}
\phi_{CF_{seq}} \defeq \quad & \fx(e_{i+1}) \Rightarrow \fx(e_i) \label{enc:cf:seq} \\
\phi_{CF_{br}}  \defeq \quad & [\fx(e_i) \Rightarrow \fx(e_k)] \land \cdots \land [\fx(e_{i+j}) \Rightarrow \fx(e_k)] \nonumber \\
                       \land & [\fx(e_i) \land \fx(e_k) \Rightarrow g_{i,k}] \land \cdots \land [\fx(e_{i+j}) \land \fx(e_k) \Rightarrow g_{i+j,k}] \nonumber \\
                       \land & \cdots \nonumber \\
                       \land & \bigvee_{e_l \in \ \texttt{succ}(e_m)} \ \
                               \bigvee_{\begin{subarray}{l}e_n \in \ \texttt{succ}(e_k) \\ e_n \neq e_m\end{subarray}} 
                               \lnot [\fx(e_m) \land \fx(e_n)]  \label{enc:cf:br} \\
\phi_{CF_{mer}} \defeq \quad & \fx(e_k) \Rightarrow (\bigvee\limits_{e_p \in \ \texttt{pred}(e_k)}^{} \fx(e_p)) \label{enc:cf:merge}
\end{align}

For these cases, we propose the encoding scheme that uniquely encodes each node of graph and at the same time allows to encode partially executed program.
%TODO: say about epsilon on (c) picture
Equation~\ref{enc:cf:seq} shows the encoding for the sequential control-flow represented in Figure~\ref{encode:cf:seq} and reflects the fact that the event $e_2$ can be executed iff the event $e_1$ has been executed.
Equation~\ref{enc:cf:br} shows the encoding for the branching control-flow depicted in Figure~\ref{encode:cf:br}, that considers both transitions and guards.
 %(similarly to the sequential control-flow)  %$\{\emptyset, \; (e_1), \; (e_1\,\Rightarrow\,e_2), \; (e_1\,\Rightarrow\,e_3)\}$.
Also, adding negations of pairwise conjunctions over all successors of the branching node, the encoding forbids the execution of two branches simultaneously.
Equation~\ref{enc:cf:merge} shows the encoding for the control-flow of a merge-point represented in Figure~\ref{encode:cf:merge}: the event $e_k$ is executed if either of its predecessors has been executed, regardless the type of the transition.
Note that the sequential control-flow is a special case of branching with the only transition guarded by $\epsilon$ (that is encoded as \texttt{true}).
%TODO: maybe add the overall formula \phi_{CF} ?

@@

Following~\cite{Porthos17b}, the indexed references of variables are computed in accordance with the following rules:
%\begin{enumerate}[noitemsep,topsep=0pt]
(1)~any access to a shared variable (both read and write) increments its SSA-index;
(2)~only writes to a local variable increment its SSA-index (reads preserve indices);
(3)~no access to a constant variable or computed (evaluated) expression changes their %SSA-index.
%\end{enumerate}
These rules determine the following encoding of load, store and computation events within single thread:
%TODO: maybe add the overall formula \phi_{DF} ?
%
\begin{align}
    \phi_{DF_{e = \texttt{load}(r \leftarrow l)}} \defeq \ & \fx(e) \Rightarrow (r_{i+1} = l_{i+1}) \\
    \phi_{DF_{e = \texttt{store}(l \leftarrow r)}} \defeq \ & \fx(e) \Rightarrow (l_{i+1} = r_i) \\
    \phi_{DF_{e = \texttt{eval}(\cdot)}} \defeq \ & \fx(e) \Rightarrow \fv(e) % \\
\end{align}

@@

The time of described algorithm is linear of the size of event-flow graph since it performs only single traverse of the graph.
%Notwithstanding the overhead of storing (or, equivalently, computing with %TODO: are you sure that 'with'?
%linear time) the predecessors of each node in order to be able to convert the program into an SSA form, the time of such a transformation reduced %TODO: How much ????
%comparing to the algorithm implemented in \porthos, which recomputed SSA-indices recursively for each instruction. %TODO: PROOF OR REMOVE !  Maybe, some words of copying, merging SSA maps ?

As it has been described before, the \rf-relation links data-flow between events %TODO: another word?
of data-flow stored in equivalence assertions over the SSA-variables. 
%The encoding of the \rf-relation, which links the SSA-variables to the original variables, 
The encoding of this linkage left untouched as it is implemented in \porthos: for each pair of events $e_1$ and $e_2$ linked by the \rf-relation, we add the following constraint:
%
\begin{align}
\phi_{DF_{mem}}(e_1, e_2) \defeq \ & \texttt{rf}(e_1, e_2) \Rightarrow (l_i = l_j)
\end{align}

>>>>

The other idea is to use satisfiability solvers for symbolic exploration of state space~\cite{clarke2001bounded}.
In this approach, the state space exploration consists of the sequence of queries to the SAT-solver, represented as boolean formulas that encode the constraints of the model and the finite path to a state in the corresponding transition system.  
%This approach uses an iterative process of constructing queries to the SAT-solver as a boolean formula which encodes the constraints of the model and the finite path to a state in the corresponding transition system. 
%Due to the SAT-solver.
This technique is called \textit{bounded model checking (BMC)} as the search process is being repeated up to the user-defined bound $k$, which may result to incomplete analysis in general case.
However, there exist numerous techniques for making BMC complete for finite-state systems~(e.g.,~\cite{shtrichman2000tuning}).
%As well as the BMC problem, the approach used by \porthos{}

@@

The control-flow of a program is represented by the \textit{control-flow graph}, a directed acyclic connected graph with a single source and multiple sink nodes. %TODO: footnote why multiple sinks
In a control-flow graph, there are two types of edges (transitions): \textit{primary transitions} that denote unconditional jumps or if-true-transitions (pictured with solid lines), and \textit{alternative transitions} that denote if-false-transitions (pictured with dotted lines).
Each node on graph can have either one successor (primary) or two successors (both primary and alternative); only computation events can serve as a branching point).
However, each merge node can have any positive number of predecessors, where each edge may be either primary or alternative.

While working on \porthos[2], we have applied some modifications to the encoding scheme for the control-flow.
The changes are conditioned by the need to be able to process an arbitrary control-flow produced by conditional and unconditional jumps of the C language.
For that, we compile the \textit{Abstract Syntax Tree~(AST)} of the parsed C-code to the plain event-flow graph.
We show %TODO! to show, actually!!!
that the new encoding is smaller than the old one used in \porthos{} since it does not produces new variables for each high-level statement of the input language.
%The \tool{PORTHOS} tool encodes the \textit{instructions}, which have recursive nature, in a recursive manner. This means, it adds synthetic composite instructions for both linear (sequential) and non-linear (branching) instructions. 
For instance, \porthos{} used the encoding scheme where the control-flow of the sequential instruction $i_1 = i_2; i_3$ was encoded as
$\phi_{CF}(i_2;i_3) \defeq (cf_{i_1} \Leftrightarrow (cf_{i_2} \land cf_{i_3})) \land \phi_{CF}(i_2) \land \phi_{CF}(i_3)$,
and control-flow of the branching instruction $i_1 = (c\,?\,i_2\!:\!i_3)$ was encoded as
$\phi_{CF}(c\,?\,i_2\!:\!i_3) \defeq (\textit{cf}_{i_1} \Leftrightarrow (\textit{cf}_{i_2} \lor \textit{cf}_{i_3})) \land \phi_{CF}(i_2) \land \phi_{CF}(i_3)$
(here we used the notation of C-like ternary operator `\texttt{x?y:z}' for defining the conditional expression `$\texttt{if}\;x\;\texttt{then}\;y\;\texttt{else}\;z$').
In contrast, the new scheme implemented in \porthos[2] firstly compiles the recursive high-level code into the linear low-level event-based representation, that is then encoded into an SMT-formula. The encoding of branching nodes depends on the \textit{guards}, the value of conditional variable on the branching state, which in turn is encoded as data-flow constraint (see Section~\ref{ch:enc:enc:df}).

@@

\begin{align}
\phi_{CF_{seq}} \defeq \quad & \fx(e_{i+1}) \Rightarrow \fx(e_i) \label{enc:cf:seq} \\
\phi_{CF_{br}}  \defeq \quad & [\fx(e_i) \Rightarrow \fx(e_k)] \ \land \ \cdots \ \land \ [\fx(e_{i+j}) \Rightarrow \fx(e_k)] \nonumber \\
                       \land & \ [\fx(e_i) \land \fx(e_k) \Rightarrow g_{i,k}] \ \land \ \cdots \ \land \ [\fx(e_{i+j}) \land \fx(e_k) \Rightarrow g_{i+j,k}] \nonumber \\
                       \land & \ \cdots \nonumber \\
                       \land & \ ( \bigvee_{e_l \in \ \texttt{succ}(e_m)} \ \
                               \bigvee_{\begin{subarray}{l}e_n \in \ \texttt{succ}(e_k) \\ e_n \neq e_m\end{subarray}} 
                              \! \lnot [\fx(e_m) \land \fx(e_n)] \ ) \label{enc:cf:br} \\
\phi_{CF_{mer}} \defeq \quad & \fx(e_k) \Rightarrow (\bigvee\limits_{e_p \in \ \texttt{pred}(e_k)}^{} \fx(e_p)) \label{enc:cf:merge}
\end{align}

For these cases, we propose the encoding scheme that uniquely encodes each node of graph and at the same time allows to encode partially executed program.
%TODO: say about epsilon on (c) picture
Equation~\ref{enc:cf:seq} shows the encoding for the sequential control-flow represented in Figure~\ref{encode:cf:seq} and reflects the fact that the event $e_2$ can be executed iff the event $e_1$ has been executed.
Equation~\ref{enc:cf:br} shows the encoding for the branching control-flow depicted in Figure~\ref{encode:cf:br}, that considers both transitions and guards.
Also, adding negations of pairwise conjunctions over all successors of the branching node, the encoding forbids the execution of two branches simultaneously.
Equation~\ref{enc:cf:merge} shows the encoding for the control-flow of a merge-point represented in Figure~\ref{encode:cf:merge}: the event $e_k$ is executed if either of its predecessors has been executed, regardless the type of the transition.
Note that the sequential control-flow is a special case of branching with the only transition guard $\epsilon$ (that is encoded as \texttt{true}).
%TODO: maybe add the overall formula \phi_{CF} ?

@@

Following~\cite{Porthos17b}, the indexed references of variables are computed in accordance with the following rules:
%\begin{enumerate}[noitemsep,topsep=0pt]
(1)~any access to a shared variable (both read and write) increments its SSA-index;
(2)~only writes to a local variable increment its SSA-index (reads preserve indices);
(3)~no access to a constant variable or computed (evaluated) expression changes their %SSA-index.
%\end{enumerate}
These rules determine the following encoding of load, store and computation events within single thread:
%TODO: maybe add the overall formula \phi_{DF} ?
%
\begin{align}
    \phi_{DF_{e = \texttt{load}(r \leftarrow l)}}  \defeq \ & [\fx(e) \Rightarrow (r_{i+1} = l_{i+1})] \\
    \phi_{DF_{e = \texttt{store}(l \leftarrow r)}} \defeq \ & [\fx(e) \Rightarrow (l_{i+1} = r_i)] \\
    \phi_{DF_{e = \texttt{eval}(\cdot)}}           \defeq \ & [\fx(e) \Rightarrow \fv(e)] % \\
\end{align}

@@

The time of described algorithm is linear of the size of event-flow graph since it performs only one graph traverse.
%Notwithstanding the overhead of storing (or, equivalently, computing with %TODO: are you sure that 'with'?
%linear time) the predecessors of each node in order to be able to convert the program into an SSA form, the time of such a transformation reduced %TODO: How much ????
%comparing to the algorithm implemented in \porthos, which recomputed SSA-indices recursively for each instruction. %TODO: PROOF OR REMOVE !  Maybe, some words of copying, merging SSA maps ?

As it has been described before, the \rf-relation links data-flow between events %TODO: another word?
of data-flow stored in equivalence assertions over the SSA-variables. 
%The encoding of the \rf-relation, which links the SSA-variables to the original variables, 
The encoding of this linkage left untouched as it is implemented in \porthos: for each pair of events $e_1$ and $e_2$ linked by the \rf-relation, we add the following constraint:
%
\begin{align}
\phi_{DF_{mem}}(e_1, e_2) \defeq \ & [\texttt{rf}(e_1, e_2) \Rightarrow (l_i = l_j)]
\end{align}

### 2018-05-18 4_impl.tex
Consider Professor's comments on abstract.

The existing implementation of \porthos{} does not distinguish the event-based program model from the high-level AST, they both are encoded into single SMT-formula (see classes of package `\texttt{dartagnan.program}' of \porthos{} tool).
Moreover, the syntax tree was implemented as a mutable data structure, which is being modified at all stages of the program (for instance, see the methods `\texttt{dartagnan.program.Program.compile(...)}' of \porthos{} that recursively compute some properties of the AST and change its state).
We are inclined to consider this architecture as one that is fast to develop, but hard to maintain (since it is difficult to guarantee the correctness of the program) and extend (since adding the support for a new high-level instruction requires changing multiple components of the program, from parser to encoder).
%Moreover, we encounter serious problems when we need to add support for control-flow jumps (as \texttt{continue}, \texttt{break}, \texttt{goto} in C).

@@

As its predecessor, \porthos[2] uses the open-source SMT solver \tool{Z3}%
%
\footnote{The \tool{Z3} project repository: \url{https://github.com/Z3Prover/z3}} %
%
by Microsoft Research~\cite{de2008z3}. However, unlike its predecessor, the \porthos[2] has an additional abstraction level, \zformula{} (see Section~\ref{ch:impl:model:zformula}), that allows the use any other SMT solver.

%We have decided to implement \porthos[2] in \textit{Java}
As its predecessor, \porthos[2] is written in \textit{Java},
%The programming language choice for \porthos[2] was also made in favour of \textit{java},
firstly, in order to be able to reuse some parts and concepts of \porthos{} also written in Java, and secondly, because the authors find the object-oriented (OOP) concepts of Java suitable for modelling languages.
Although Java does not show the best results in performance benchmarks (for example, compared to C++~\cite{hundt2011loop}), the performance cornerstone of \porthos[2] (as well as any other SMT-based code analyser) is the phase of solving the SMT-formula, which is left to the third-party SMT solver invoked by \porthos[2] via a Java API.
However, considering the perspective of using \porthos[2] as a static analyser for real-world programs, the memory optimisation problem must also be taken into account during both encoding and solving stages.
It is worth noting that, for the reasons of simplicity, the \porthos[2] is not a concurrent program, however, we believe that, due to its modular architecture, it can be easily parallelised on the level of program modules.

@@

All interface methods can be divided into two groups.
The first group constitute methods that emit events.
These methods construct an event object (as events contain the \texttt{XEventInfo} structure that stores information about the owning process, all events must be created by the process constructor, i.e. X-interpreter) and change the state of interpreter considering the newly created event.
Note that the methods \texttt{createComputationEvent} do not emit a computation event but only create one.
This is performed as optimisation that removes unused computations from the model (otherwise, for instance, the assignment 
`\texttt{x = 1 * (2 + 3)}' would be compiled to two consequent events `\texttt{eval(2 + 3); write(register <- eval(1 * eval(2 + 3)));}' instead of single event `\texttt{write(register <- eval(1 * eval(2 + 3)));}').
If the computation event was not used at all (for example, in the following C code: `\texttt{foo(); 1; bar();}' the execution event `\texttt{eval(1)}' is skipped by the model), it is also removed from the event-graph (see justification in Section~\ref{ch:impl:model:xgraph:evt}).
The second group of X-interpreter methods consists of the methods for defining non-linear statements (branchings and loops).
These methods change the state of the interpreter and set up additional non-linear control-flow edges.
%`\texttt{register = 1 * (location + 2)}' would be compiled to three consequent events `\texttt{load(temp\_reg\_1 <- location); eval(temp\_reg\_1 + 2); write(register <- 1 * (temp\_reg\_1 + 2));}',

>>>>

The existing implementation of \porthos{} does not distinguish the event-based program model from the high-level AST, they both are encoded into a single SMT-formula (see classes of package `\texttt{dartagnan.program}' of \porthos{} tool).
Moreover, the syntax tree was implemented as a mutable data structure, which is being modified at all stages of the program (for instance, see the methods `\texttt{dartagnan.program.Program.compile(...)}' of \porthos{} that recursively compute some properties of the AST and change its state).
We are inclined to consider this architecture as one that is fast to develop, but hard to maintain (since it is difficult to guarantee the correctness of the program) and extend (since adding the support for a new high-level instruction requires changing multiple components of the program, from parser to encoder).
%Moreover, we encounter serious problems when we need to add support for control-flow jumps (as \texttt{continue}, \texttt{break}, \texttt{goto} in C).

@@

As its predecessor, \porthos[2] uses the open-source SMT solver \tool{Z3}%
%
\footnote{The \tool{Z3} project repository: \url{https://github.com/Z3Prover/z3}} %
%
by Microsoft Research~\cite{de2008z3}. However, unlike its predecessor, \porthos[2] has an additional abstraction level, \zformula{} (see Section~\ref{ch:impl:model:zformula}), that allows the use any other SMT solver.

%We have decided to implement \porthos[2] in \textit{Java}
As its predecessor, \porthos[2] is written in \textit{Java},
%The programming language choice for \porthos[2] was also made in favour of \textit{java},
firstly, in order to be able to reuse some parts and concepts of \porthos{} also written in Java, and secondly, because the authors find the object-oriented (OOP) concepts of Java suitable for modelling languages.
Although Java does not show the best results in performance benchmarks (for example, compared to C++~\cite{hundt2011loop}), the performance cornerstone of \porthos[2] (as well as any other SMT-based code analyser) is the phase of solving the SMT-formula, which is left to the third-party SMT solver invoked by \porthos[2] via a Java API.
However, considering the perspective of using \porthos[2] as a static analyser for real-world programs, the memory optimisation problem must also be taken into account during both encoding and solving stages.
It is worth noting that, for the reasons of simplicity, \porthos[2] is not a concurrent program, however, we believe that, due to its modular architecture, it can be easily parallelised on the level of program modules.

@@

All interface methods can be divided into two groups.
The first group constitute methods that emit events.
These methods construct an event object (as events contain the \texttt{XEventInfo} structure that stores information about the owning process, all events must be created by the process constructor, i.e. X-interpreter) and change the state of interpreter considering the newly created event.
Note that the methods \texttt{createComputationEvent} do not emit a computation event but only create one.
This is performed as optimisation that removes unused computations from the model (otherwise, for instance, the assignment 
`\texttt{x = 1 * (2 + 3)}' would be compiled to two consequent events `\texttt{eval(2 + 3); write(register <- eval(1 * eval(2 + 3)));}' instead of a single event `\texttt{write(register <- eval(1 * eval(2 + 3)));}').
If the computation event was not used at all (for example, in the following C code: `\texttt{foo(); 1; bar();}' the execution event `\texttt{eval(1)}' is skipped by the model), it is also removed from the event-graph (see justification in Section~\ref{ch:impl:model:xgraph:evt}).
The second group of X-interpreter methods consists of the methods for defining non-linear statements (branchings and loops).
These methods change the state of the interpreter and set up additional non-linear control-flow edges.
%`\texttt{register = 1 * (location + 2)}' would be compiled to three consequent events `\texttt{load(temp\_reg\_1 <- location); eval(temp\_reg\_1 + 2); write(register <- 1 * (temp\_reg\_1 + 2));}',
